Union and intersection contracts: theory vs practice

### 2021-04-06 paper.tex
Add some bib on contracts, rewording

The explorative work presented in this paper is motived by the development of
the Nickel language\cite{NickelRepo}. Nickel is a configuration programming
language, meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight, easy to write and easy to embed in a larger system, to serve either
directly as a native language to specify configuration, or as a meta-language to
generate them. The focus is first and foremost practical: Nickel originated as a
rethink of the language of the Nix package manager, used to describe one of the
biggest sets of software packages\cite{repology}.  Nickel targets build systems
and cloud deployment tools (Kubernetes, Terraform or NixOps) as well. While
there's no hesitation in incorporating recent research when justified, the
design decisions are guided by the industrial use cases.

@@

Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
Unlike JSON, though, Nickel anticipates large configurations by being both
programmable and typed. In the configuration setting, there is a singular
dilemma with respect to typing: since a program is run on fixed inputs and is
expected to terminate, any relevant type error will show up at evaluation.  Why
bother with the complexity of a static type system? On the other hand, more and
more software systems offload complexity to configurations, such as the
infrastructure as code paradigm prevailing today in cloud deployment. When the
complexity of a codebase grows, static types becomes attractive again.  For
library code — that is, library functions —, static types are specifically
adapated and bring all the usual benefits.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
which mixes both static and dynamic typing. Unlike gradually-typed languages
like TypeScript, we do not seek to build a complex type system that tries hard
to accept most of the idioms naturally arising in dynamically typed code.
Rather, we choose to provide a reasonably expressive type system with good
inference properties, that make typing generic functions operating on data easy,
but may require using untyped code for more exotic expressions.  In any case,
this is not a surrender, as Nickel provides a complementary mechanism for more
advanced data validation: contracts.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts.  In this setting, each type constructor of the
static type system - arrows, records, foralls - must have a contract counterpart
that check at run-time that a value is indeed a member of the corresponding
type. Various type constructors have been investigated [All refs], with
solutions of varying complexity. Amongst all the common extensions to the
polymorphic lambda calculus, one of the most simple and ubiquitous turns out to
be one of the most involved: \textit{union types}.

>>>>

The explorative work presented in this paper is motived by the development of
the Nickel language\cite{NickelRepo}. Nickel is a configuration programming
language, meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight, easy to write and easy to embed in a larger system, to serve either
directly as a native language to specify configuration, or as a meta-language to
generate them. The focus is first and foremost practical: Nickel originated as a
rethink of the language of the Nix package manager, used to describe one of the
biggest sets of software packages\cite{repology}.  Nickel targets build systems
and cloud deployment tools (Kubernetes, Terraform or NixOps) as well. While
there is no hesitation in incorporating recent research or doing it ourselves,
the design decisions are always guided by the practical industrial use cases.

@@

Our aim is that writing simple Nickel feels as easy as writing JSON or YAML.
Unlike JSON, though, Nickel anticipates large configurations by being both
programmable and typed. In the configuration setting, there is a singular
dilemma with respect to typing: since a program is run on fixed inputs and is
expected to terminate, any relevant type error will show up at evaluation.  Why
bother with the complexity of a static type system? On the other hand, more and
more software systems offload complexity to configurations, such as the
infrastructure as code paradigm prevailing today in cloud deployment. When the
complexity of a codebase grows, static types becomes attractive again.  For
reusable code — that is, library functions —, static types are specifically
adapated, and bring all the usual benefits of early error detection, code
robustness, better code structure, documentation, and so on.

This dilemma is naturally solved by gradual typing\cite{Siek06gradualtyping}
which mixes both static and dynamic typing. Unlike gradually-typed languages
like TypeScript, we do not seek to build a complex type system that tries hard
to accept most of the idioms naturally arising in dynamically typed code.
Rather, we choose to provide a reasonably expressive type system with good
inference properties, that make typing functions operating on generic data easy,
but may require using untyped code for more exotic expressions.  In any case,
this is not a surrender, as Nickel provides a complementary mechanism for more
advanced data validation: contracts.

@@

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Fellleisen introduces a principled approach to run-time assertion checking that
nicely support higher-order functions and introduce the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}. Nickel adopts
a similar type system, with both statically typed terms, dynamically typed
terms, and first-class contracts.  In this setting, each type constructor of the
static type system - arrows, records, foralls - must have a contract counterpart
that check at run-time that a value is indeed a member of the corresponding
type. Contracts for various extensions of the simply typed $\lambda$-caluclus
have been investigated \cite{BlameForAll, KeilThiemannUnionIntersection,
RootCauseOfBlame, DependentContracts, GradualTypingClasses}, with solutions of
varying complexity.  Amongst all the common extensions to the polymorphic lambda
calculus, one of the most simple and ubiquitous turns out to be one of the most
involved: \textit{union types}.
